---
title: "Data Wrangling 2"
subtitle: "Complex sequences of manipulations"
author: Diogo Melo
date: Oct. 4th, 2023
runtime: static
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, "styles.css"]
    nature:
      highlightLines: true
      chakra: libs/remark-latest.min.js
      beforeInit: "macros.js"
      ratio: "16:10"
      countdown: 60000
---


class: inverse, center, middle
# Processing data sets 

---

# Counting along columns

- The most basic processing we can do is counting the number of rows in a column that share the same value 

```{r}
flights |> 
  count(carrier)
```

---

# Grouping using `group_by`

- We can also separate our processing using some grouping variable
- All sensible operations happen inside groups in a grouped data.frame
 

```{r}
flights %>%
  group_by(carrier) %>%
  count(origin)
```

---

# Summarizing values using `summarize`

.pull-left[
- `summarize` can be used to perform operations in the columns

```{r}
flights %>%
  summarise(mean_delay = mean(arr_delay))
```
]
--

.pull-right[
- With `group_by`, this is a powerful function:

```{r}
flights %>%
  group_by(carrier) %>%
  summarize(delay = mean(arr_delay))
```
]

# Combining data sets 

- Sometimes the information we want is split across two data.frames

- Maybe we have two types of measures, or summaries that were calculated separately

---

# Match

---


# col and row bind

---


# Datatable alternatives

---