---
title: "Data visualization"
subtitle: "Bulding pretty plots in ggplot2" 
author: Diogo Melo
date: Oct. 30th, 2023
runtime: static
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts, "styles.css"]
    nature:
      highlightLines: true
      chakra: libs/remark-latest.min.js
      beforeInit: "macros.js"
      ratio: "16:10"
      countdown: 60000
---

# Chosing plot types

---

# Color scales

---

# Using mappings to convey information

---

# Scales

- Scales define how the mapping you specify inside `aes()` should happen. All 
mappings have an associated scale even if not specified.

```{r, eval = FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class))
```

 - We can add one explicitly. All scales follow the same naming conventions.

```{r, eval = FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, colour = class)) + 
  scale_colour_brewer(type = 'qual')
```

- Positional mappings (x and y) also have associated scales.

```{r, eval = FALSE}
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) + 
  scale_x_continuous(breaks = c(3, 5, 6)) + 
  scale_y_continuous(trans = 'log10')
```

---

# Themes

---

# Text and fonts

---

# Annontations vs legends

---