---
title: "EEB330 -- Precept 01: Control flow"
subtitle: "Simple control flow in R"
author: Scott Wolf
date: Sept. 20th, 2023
output: html_document
---

# Introduction to control flows

Control flows in R determine the sequence in which the instructions are executed. This ability to control the flow of execution allows for more complex and flexible scripts. 

## Basic conditionals

In R, the basic conditional statements are `if`, `else`, and `else if`.

```{r}
set.seed(42) # This ensures that the random number generator will produce the same results each time the script is run
x <- runif(1, 0, 10)

if (x > 5) {
     print("x is greater than 5")
} else if (x < 5) {
     print("x is less than 5")
} else {
     print("x is equal to 5")
}
```

## Loops

There are two primary loops in R - `for` and `while`.

**For loop**

```{r}
for (i in 1:5) {
     print(i)
}
```

**While loop**

```{r}
counter <- 1

while (counter <= 5) {
     print(counter)
     counter <- counter + 1
}
```

## Functions

Functions are blocks of reusable code.

```{r}
calculate_sum <- function(a, b) {
     return(a + b)
}

calculate_sum(3, 4)
```

### Scoping

Scoping refers to the visibility of variables. In R, there are two main scopes: global and local.

```{r}
global_var <- 10

function_with_scope <- function() {
  local_var <- 5
  return(global_var + local_var)
}

function_with_scope()
```

### Recursion

A function that calls itself is known as a recursive function.

```{r}
factorial <- function(n) {
     if (n <= 1) {
          return(1)
     } else {
     return(n * factorial(n-1))
     }
}

factorial(5)
```

# Submission

For this precept, please submit a single R Markdown file named `precept01.Rmd`. Please make intermediate commits in your development branch as you write to get in the habit of committing often. You should commit at least once for each question. Finally, you should submit a pull request to merge your development branch into your master branch. As a reminder, once you merge this PR, your assignment will be marked as complete and may be graded.


## Exercises


### 1. Conditionals

Consider the vector:

```{r}
grades <- c(85, 77, 90, 65, 89, 72, 91, 68, 80)
```

Write a conditional script that:

1. Tags grades above 90 as "Excellent"
2. Between 80 and 90 (inclusive) as "Good"
3. Below 80 as "Fair"

Store the result in a new vector called grade_tags. Ensure the order in grade_tags corresponds to the order of grades in the grades vector.


### 2. For

Create a sequence of numbers from 3 to 15. Using a for loop, compute the cumulative product of these numbers.


### 3. While

Starting with a number 50, decrease it by 5% in every iteration using a while loop. Continue the iterations until the number goes below 20. Return the number of iterations it took.


### Functions with conditionals and loops

Write a function named series_sum that accepts two arguments: a start value 's' and an end value 'e'. The function should sum all numbers from 's' to 'e'. However, if a number in the series is divisible by 3, it should be skipped.


### 5. Recursive Functions

The Lucas series is a sequence of numbers similar to the Fibonacci series but starts with 2 and 1 instead of 0 and 1. Write a recursive function lucas that computes the nth value of the Lucas series.