---
title: "EEB330 -- Precept 06: Plotting"
author: Scott Wolf, Michelle White
date: Oct. 24, 2024
output: html_document
---

## [GitHub Classroom Assignment](https://classroom.github.com/a/kCg_EfG5)
- Check out the cheat sheet for ggplot2: [Data Visualization with ggplot2: CHEAT SHEET](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
- Similarly for matplotlib: [Matplotlib for Beginners: CHEAT SHEET](https://matplotlib.org/cheatsheets/_images/handout-beginner.png)
- *Choose one language (R or Python) to complete the exercises, but complete the first exercise in the opposite language. Submit separate files for each language.*
- *Deadline is back to one week from the original precept date.*
- *Please save each plot as an image (.png or .jpg) and upload them to your GitHub repository (4 images total).*

```{r, echo = T, message = F, results = "hide", warning = F}
# R
library(pak)
packages <- c("ggplot2", "cowplot", "ggthemes", "patchwork")
pak::pkg_install(packages)

lapply(packages, require, character.only = TRUE)
```

```{python, echo = T, message = F}
# Python
import matplotlib.pyplot as plt
import seaborn as sns
import pandas as pd

mpg = pd.read_csv("mpg.csv")
```

## Exercises

### Exercise 1 -- Bar Plot Modification

- **Task:**
   * Add a title to the plot: "Distribution of Cars by Class".
   * Change the x-axis label to "Type of Car".
   * Color the bars in blue.
   * Rotate the x-axis labels by 45 degrees.

- **Expected Output:** An updated plot with the above specifications.

Initial Plot: A simple bar plot displaying the number of cars for each class in the mpg dataset.
```{r, fig.show='hide'}
# R
ggplot(data = mpg, aes(x = class)) +
  geom_bar()
```

```{python, fig.show='hide'}
# Python
sorted_mpg = mpg['class'].value_counts().sort_index()
plt.bar(sorted_mpg.index, sorted_mpg.values)
```


### Exercise 2 -- Histogram Modification

- **Task:** 
  * Add a title to the plot: "Highway Mileage Distribution".
  * Change the x-axis label to "Miles Per Gallon".
  * Fill the histogram bars with green but have a black border.
  * Set the bin width to 2.

- **Expected Output:** An updated plot with the above specifications.

Initial Plot: A histogram showcasing the distribution of highway miles per gallon (hwy) from the mpg dataset.
```{r, fig.show='hide'}
# R
ggplot(data = mpg, aes(x = hwy)) + 
  geom_histogram()
```

```{python, fig.show='hide'}
# Python
plt.hist(mpg['hwy'], bins = 30)
```


### Exercise 3 -- Scatter Plot with Facets

- **Task:**
   * Add a title: "Engine Displacement vs. Highway MPG".
   * Change the x-axis label to "Engine Size (liters)" and y-axis label to "Highway MPG".
   * Color the points by class and shape them by the type of drive (e.g., 4wd, fwd, rwd).
   * Add a smooth trend line (with standard error or confidence interval) to the plot. Consider adjusting the alpha of the points for clarity.
   * Facet the plot by cyl (number of cylinders) in a 2x2 grid format.

- **Expected Output:** An updated plot with the above specifications.

Initial Plot: A scatter plot illustrating the relationship between engine displacement (displ) and highway MPG (hwy).
```{r, fig.show='hide'}
# R
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```

```{python, fig.show='hide'}
# Python
sns.scatterplot(data = mpg, x = 'displ', y = 'hwy')
```

### Exercise 4: Enhanced Boxplots using after_stat() and patchwork

- **Task:** 
1. Modify plot1:
   * Color the boxes based on median value of cty using a gradient from light blue (low mpg) to dark blue (high mpg).
   * Add a title: "City MPG by Manufacturer".
   * Rotate x-axis labels by 90 degrees and adjust their size for readability.
   * Apply a theme of your choice from the ggthemes library (R) or `.set_theme()` in the Seaborn library (Python).

2. Modify plot2:
   * Color the boxes based on median value of hwy using a gradient from light green (low mpg) to dark green (high mpg).
   * Add a title: "Highway MPG by Manufacturer".
   * Rotate x-axis labels by 90 degrees and adjust their size for readability.
   * Apply the same theme as plot1.

3. Combine the two modified plots side by side using the patchwork library (R) or `subplots` (Python).

- **Expected Output:** A plot with the above specifications.

```{r, fig.show='hide'}
# R
plot1 <- ggplot(data = mpg, aes(x = manufacturer, y = cty)) +
  geom_boxplot()

plot2 <- ggplot(data = mpg, aes(x = manufacturer, y = hwy)) +
  geom_boxplot()
```

```{python, fig.show='hide'}
# Python
plot1 = sns.boxplot(data = mpg, x = 'manufacturer', y = 'cty')

plot2 = sns.boxplot(data = mpg, x = 'manufacturer', y = 'hwy')
```
